He-Man: A domain-specific language for lightweight concurrency

### 2012-03-11 paper/paper.tex
Finished related work, conclusion.

% XXX: standardize on IO vs I/O
% XXX: standardize on continuation-passing vs continuation passing
% XXX: standardize on domain-specific vs domain specific
% TODO: talk about type system, runtime improvements, optimizations

%XXX: new title?
\title{He-Man: A domain specific language for lightweight concurrency}
%\subtitle{I said hey, what's goin' on}
\authorinfo{Carlo Angiuli}{Carnegie Mellon University}{cangiuli@cs.cmu.edu}
\authorinfo{Michael Sullivan}{Carnegie Mellon University}{mjsulliv@cs.cmu.edu}

@@

This paper describes the design and Haskell implementation of He-Man, a domain
specific language for network application programming which presents a
sequential thread-like abstraction, but compiles to performant event-driven C
code. Our implementation is available at
\url{https://github.com/msullivan/he-man}. %\cite{HeMan}
\end{abstract}

@@

The idea of implementing custom schedulers for lightweight thread abstractions
is not new, and has been used to great effect in functional language runtimes
like GHC \cite{MarlowEtAl}. Li and Zdancewic \cite{LiZdancewic} have developed
monadic abstractions in Haskell for threaded and event-driven concurrency, for
the same problem domain of scalable network applications; their implementation
draws upon the concurrency and synchronization support in the GHC runtime.

@@

Our contribution is to distill the Lauer-Needham duality to its essence, by
compiling from a simple thread abstraction to self-contained event-driven C
code. This abstraction, He-Man: Haskell Event Manager, Apropos Networking, is a
compiled domain specific language embedded in Haskell, which offers a small
monadic language supporting thread spawning, blocking events, and external
function calls.

@@

The underlying abstract syntax for the front-end He-Man language is a
fairly conventional imperative, block-structured, statement-based,
first-order language, shown in Figure \ref{fig:front-syntax}. The
notable constructs are \b{Spawn}, creating a new thread, which takes a
list of arguments and launches the associated code with those
arguments available to it, and \b{Wait}, which takes an event
expression as an argument and blocks until the event is
``ready''. Note that \b{Call} is for calling foreign functions; He-Man
does not have its own notion of function.

@@

-- Some typed He-Man functions
(.<) :: IntE -> IntE -> BoolE
(.<) = typ2 (RelnOp Lt)
(.==) :: Expr a -> Expr a -> BoolE
(.==) = typ2 (RelnOp Eq)
var :: String -> Type a -> Expr a -> Prog (Expr a)
var = ...
\end{code}
\caption{He-Man expression typing} % XXX
\label{fig:types}
\end{figure}

@@

\end{code}
\caption{He-Man argument packets} % XXX
\label{fig:arg-packets}
\end{figure}

@@

Our back-end's main task is to convert our linear front-end code into a state
machine of code blocks. This occurs in our \t{flattenPrgm} pass, which passes
through program \i{Stmt}s in reverse order, generating uniquely-named blocks
whose only control-flow consists of (non-nested) tail-position jumps to other
blocks. Threads are identified with their first block.
% TODO discuss the transformations

@@

Once our 
% TODO fuseBlocks pass

@@

Currently, the He-Man runtime provides support for two types of
events: socket/pipe file I/O and asynchronous He-Man channels.  The
runtime has partial support for Linux Kernel Asynchronous IO for file
IO, using a Linux \t{eventfd} to allow \t{epoll} to wait on AIO
events. Adding new types of events that an application requires would
be simple, but require some runtime hacking.

@@

The He-Man version is 33 lines long and clearly represents the control flow of
each individual thread. (This does not include the He-Man library, which has 25
lines of code implementing generic routines for simulating blocking IO using
nonblocking IO and wait.) The total line counts, somewhat understate the
difference between the versions, since the boilerplate for setting up
connections is about the same for both versions. When just looking at the core
read/write loop, the difference is much more apparent: the C version is 38 lines
of code, and the He-Man version is 5.

@@

\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_100.pdf}
\caption{Web server benchmark, 100 byte file.}
\label{fig:graph100}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_16K.pdf}
\caption{Web server benchmark, 16 KB file.}
\label{fig:graph16}
\end{figure*}
\begin{figure*}[!ht]
\centering
\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
\caption{Web server benchmark, 128 KB file.}
\label{fig:graph128}
\end{figure*}

@@

% XXX: remains unnecessarily general for us?

He-Man is most similar to Li and Zdancewic's Haskell implementation of
a hybrid event/thread programming model for network applications
\cite{LiZdancewic}.  Their project has similar goals to ours, but is
implemented entirely within Haskell. As a result, unlike He-Man, their
framework requires entire applications to be written in Haskell, and
incurs runtime overhead from the Haskell system.

While He-Man's goals are similar to
those of several other systems, we feel that it solves

\subsection{Lightweight threads}

% Related work:
%\cite{VanderwaartCrary} re bounded time
%\cite{vonBehrenConditBrewer} re cooperative multithreading
%\cite{LiZdancewic}

% Haskell EDSLs:
%\cite{Gill}
%\cite{ElliottEtAl} CITE THIS
%\cite{LeijenMeijer}

% Functional programming:
\cite{Vinoski}
\cite{Hellstrom}
FoxNet for Standard ML \cite{BiagioniEtAl}
\cite{GraunkeEtAl}
\cite{Queinnec}

The C10K problem is particularly evident in the context of HTTP
servers, which routinely maintain many concurrent connections. As we
discussed in Section \ref{sec:evaluation}, Apache does not scale
particularly well. The popular Lighttpd and Nginx HTTP server projects
were both started with the explicit goal of outperforming Apache with
respect to the C10K problem \cite{Lighttpd,Nginx}, and both use
event-based architectures similar to He-Man.

%A number of custom network stack implementations exist for functional
%languages, such as FoxNet for Standard ML \cite{BiagioniEtAl}, and

There exist C/C++ libraries for event loops\cite{libevent}, but none of these
avoid the state machine paradigm.

%% XXX: needs serious work
\section{Future work}

We also plan to expand the language with the ability to declare
functions and make non-external function calls within a thread. This
will make the generated code smaller by allowing the programmer to
abstract away common functionality like read and write loops.

The runtime could be much more carefully tuned.
Because the runtime allocates a \t{struct} on every thread spawn, we should
replace our \t{malloc} calls with calls to a custom slab allocator.

We also hope to implement observable sharing, as in \cite{Gill}, to allow the
same thread to have distinct spawn sites without code duplication. This does not
add inefficiency, but does increase the size of our output code.

\section{Conclusions}
Expressing network application logic as a compiled, embedded, domain
specific language is an elegant and practical way to build scalable
network applications without having to manually write state-machine
based event-loop code. Embedding the language in Haskell allows
powerful metaprogramming and allowed the language to be developed with
a fraction of the effort of building a full language. Compiling to C
allows C programmers to write parts of their system (for example, the
caching layer) in C while writing the network logic in
straightforward, straight-line He-Man code.

% XXX

We identify our contributions as:
* We present the idea of generating code for network facing logic from a
description written in a DSL while building the rest of the application in
C. In a way, the DSL could be considered as a method for programmatically
generating event driven frameworks customized for the application.
* We observe that converting a traditional block-structured imperative
program to a form amenable for use in an event-driven system is analogous
to CPS conversion and thus easily done automatically. (Is this actually a
contribution, or is it obvious and been said a million times before?)
* We present a case study of compiled embedded domain specific language
design and implementation
* We demonstrate that it is practical to build efficient network
applications using our system

\cite{*}

>>>>

%There exist C/C++ libraries for event loops\cite{libevent}, but none of these
%avoid the state machine paradigm.

%XXX: new title?
\title{He-Man: A domain-specific language for lightweight concurrency}
%\subtitle{I said hey, what's goin' on}
\authorinfo{Carlo Angiuli}{Carnegie Mellon University}{cangiuli@cs.cmu.edu}
\authorinfo{Michael Sullivan}{Carnegie Mellon University}{mjsulliv@cs.cmu.edu}

@@

This paper describes the design and Haskell implementation of He-Man, a
domain-specific language for network application programming which presents a
sequential thread-like abstraction, but compiles to performant event-driven C
code. Our implementation is available at
\url{https://github.com/msullivan/he-man}. %\cite{HeMan}
\end{abstract}

@@

The idea of implementing custom schedulers for lightweight thread abstractions
is not new, and has been used to great effect in functional language runtimes
like GHC \cite{MarlowEtAl}. Li and Zdancewic \cite{LiZdancewic} have developed
monadic abstractions in Haskell for threaded and event-driven concurrency, for
the same problem domain of scalable network applications; their implementation
draws upon concurrency and synchronization support in the GHC runtime.

@@

Our contribution is to distill the Lauer-Needham duality to its essence, by
compiling from a simple thread abstraction to self-contained event-driven C
code. This abstraction, He-Man: Haskell Event Manager, Apropos Networking, is a
compiled domain-specific language embedded in Haskell, which offers a small
monadic language supporting thread spawning, blocking events, and external
function calls.

@@

The underlying abstract syntax for the front-end He-Man language is a
fairly conventional imperative, block-structured, statement-based,
first-order language, shown in Figure \ref{fig:front-syntax}. The
notable constructs are \b{Spawn}, creating a new thread, which takes a
list of arguments and launches the associated code with those
arguments available to it, and \b{Wait}, which takes an event
expression as an argument and blocks until the event is
``ready''. Note that \b{Call} is for calling foreign functions; He-Man
does not have its own notion of function.\footnote{Adding functions or
observable sharing to He-Man could reduce output code size, but would have
little impact on performance.}

@@

-- Some typed He-Man functions
(.<) :: IntE -> IntE -> BoolE
(.<) = typ2 (RelnOp Lt)
(.==) :: Expr a -> Expr a -> BoolE
(.==) = typ2 (RelnOp Eq)
var :: String -> Type a -> Expr a -> Prog (Expr a)
var = ...
\end{code}
\caption{Implementation of He-Man type system.}
\label{fig:types}
\end{figure}

@@

\end{code}
\caption{Implementation of He-Man \t{ArgPacket}s.}
\label{fig:arg-packets}
\end{figure}

@@

Our back-end's main task is to convert our linear front-end code into a state
machine of code blocks. This occurs in our \t{flattenPrgm} pass, which passes
through program \i{Stmt}s in reverse order, generating uniquely-named blocks
whose only control-flow consists of (non-nested) tail-position jumps to other
blocks. Threads are identified with their first block.

TODO here we will discuss the transformations that occur

@@

TODO here we discuss the fuseBlocks pass

@@

Currently, the He-Man runtime provides support for two types of
events: socket/pipe file I/O and asynchronous He-Man channels.  The
runtime has partial support for Linux Kernel Asynchronous I/O for file
I/O, using a Linux \t{eventfd} to allow \t{epoll} to wait on asynchronous I/O
events. Adding new types of events that an application requires would
be simple, but require some runtime hacking.

@@

The He-Man version is 33 lines long and clearly represents the control flow of
each individual thread. (This does not include the He-Man library, which has 25
lines of code implementing generic routines for simulating blocking I/O using
nonblocking I/O and wait.) The total line counts, somewhat understate the
difference between the versions, since the boilerplate for setting up
connections is about the same for both versions. When just looking at the core
read/write loop, the difference is much more apparent: the C version is 38 lines
of code, and the He-Man version is 5.

@@

\begin{figure*}[htbp]
\centering
\includegraphics[width=0.9\textwidth]{graph_100.pdf}
\caption{Web server benchmark, 100 byte file.}
\label{fig:graph100}
\end{figure*}
\begin{figure*}[htbp]
\centering
\includegraphics[width=0.9\textwidth]{graph_16K.pdf}
\caption{Web server benchmark, 16 KB file.}
\label{fig:graph16}
\end{figure*}
\begin{figure*}[htbp]
\centering
\includegraphics[width=0.9\textwidth]{graph_128K.pdf}
\caption{Web server benchmark, 128 KB file.}
\label{fig:graph128}
\end{figure*}

@@

Li and Zdancewic \cite{LiZdancewic} have developed a hybrid concurrency monad in
Haskell for building scalable network applications. Their implementation uses
the CPS monad to represent computations as continuations in a ready queue, and
gives the illusion that I/O is blocking in each thread, while it is actually
handled with foreign calls to \t{epoll}. This concurrency monad is executed in
Haskell, unlike He-Man programs, which are compiled for speed and portability
\cite{ElliottEtAl}.

\subsection{Functional network programming}

In addition to the concurrency models mentioned above, a number of functional
languages have network libraries specifically tailored to concurrency issues,
like Erlang's TCP/IP suite \cite{ParisEtAl}, and the FoxNet library for Standard
ML \cite{BiagioniEtAl}, which uses continuations to implement coroutines for
cooperative multithreading.

It is important to note that in environments providing cooperative
multithreading, like Capriccio, FoxNet, and He-Man, while the runtime guarantees
that no thread monopolizes execution while waiting for I/O, it is certainly
possible for one thread to monopolize execution by running indefinitely.
Bounding execution time is a difficult problem which has been studied in the
context of grid computing \cite{VanderwaartCrary}. The He-Man compiler
guarantees that no code block contains an infinite loop; unbounded execution can
only occur in calls to external functions.

The Scheme community often uses continuations to develop web applications which
robustly handle client state despite the statelessness of HTTP. This can be
accomplished either by CPSing a direct-style program \cite{GraunkeEtAl} or by
modifying a HTTP server to explicitly store continuations \cite{Queinnec}.

\section{Conclusion}

The Lauer-Needham duality suggests that the performance of event-driven
concurrent code should be achievable with simpler threaded semantics. We propose
a model of network application programming which completely separates I/O logic,
which requires a tuned concurrent implementation, from largely single-threaded
protocol and implementation details.

We present He-Man, a domain-specific language for expressing network application
I/O in a straightforward threaded abstraction, which compiles to efficient
event-driven C code. We demonstrate that it is practical to build efficient
network applications with a combination of He-Man and C code, and unlike other
hybrid concurrency systems, the He-Man runtime is very compact and easily
modified.
